---
title: "MATH 680: Assignment 2"
author: "Annik Gougeon, David Fleischer"
date: "Last Update: `r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
header-includes:
   - \usepackage{amsmath,amsthm,amssymb,mathtools}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\R}{\mathbb R}
\renewcommand{\S}{\mathbb S}

# Question 1

\begin{proof} Let $C = \left\{ x \in \R^n: Ax \leq b \right\}$ be our set of interest. Let $x, y \in C$, and let $t \in [0, 1]$ be an arbitrary real-valued scalar. Then,
\begin{align*}
  A \left( tx + (1 - t)y \right) &= tAx + (1 - t)Ay \\
  &\leq tb + (1 - tb) \\
  &= b.
\end{align*}

Thus,
$$
  x, y \in C \implies tx + (1 - t)y \in C, \quad \text{for all } 0 \leq t \leq 1.
$$
That is, $C$ is a convex set, as desired.
\end{proof}

# Question 2

## 2.1

\begin{proof} Recall that a (continuous, twice differentiable) function $f(z)$, $z\in C$, is convex on $C$ if and only if its Hessian is positive semidefinite for all $z$ on the interior of $C$,
$$
  \nabla^2 f(z) \in \S^n_+,
$$
and strongly convex with parameter $m > 0$ if and only if
$$
  \nabla^2 f(z) - m \mathbb I_n \in \S^n_{+}.
$$

Furthermore, a matrix $M \in \R^{n\times n}$ is positive semidefinite if and only if all eigenvalues of $M$ are nonnegative. Since $f$ is nondifferentiable along $x = 0, y = 0$ we first apply a differentiable approximation $f_\epsilon$
$$
  f_\epsilon(x, y) = \sqrt{x^2 y^2 + \epsilon} + a\left(x^2 + y^2\right) \underset{\epsilon\to 0}{\longrightarrow} |xy| + a\left(x^2 + y^2\right) = f(x, y)
$$
Now, $f_\epsilon$ admits gradient 
$$
  \nabla f_\epsilon(x, y) = \left(2 a x+\frac{x y^2}{\sqrt{x^2 y^2+\epsilon }},2 a y+\frac{x^2 y}{\sqrt{x^2 y^2+\epsilon }}\right),
$$

and Hessian
$$
  \nabla^2 f_\epsilon(x, y) = \left[
\begin{array}{cc}
 -\frac{x^2 y^4}{\left(x^2 y^2+\epsilon \right)^{3/2}}+\frac{y^2}{\sqrt{x^2 y^2+\epsilon }}+2 a & \frac{2 x y}{\sqrt{x^2 y^2+\epsilon }}-\frac{x^3 y^3}{\left(x^2 y^2+\epsilon \right)^{3/2}} \\
 \frac{2 x y}{\sqrt{x^2 y^2+\epsilon }}-\frac{x^3 y^3}{\left(x^2 y^2+\epsilon \right)^{3/2}} & -\frac{y^2 x^4}{\left(x^2 y^2+\epsilon \right)^{3/2}}+\frac{x^2}{\sqrt{x^2 y^2+\epsilon }}+2 a \\
\end{array}
\right].
$$

We find $\nabla^2 f_\epsilon(x, y)$ to be\footnote{Proof left as an exercise.}
\begin{align*}
  \lambda_{\epsilon, 1} &= \frac{x^2 \left(4 a y^2 \sqrt{x^2 y^2+\epsilon }+\epsilon \right)+4 a \epsilon  \sqrt{x^2 y^2+\epsilon }-\sqrt{4 x^6 y^6+x^4 \epsilon  \left(16 y^4+\epsilon \right)+14 x^2 y^2 \epsilon ^2+y^4 \epsilon ^2}+y^2 \epsilon }{2 \left(x^2 y^2+\epsilon \right)^{3/2}} \\
  \lambda_{\epsilon, 2} &= \frac{x^2 \left(4 a y^2 \sqrt{x^2 y^2+\epsilon }+\epsilon \right)+4 a \epsilon  \sqrt{x^2 y^2+\epsilon }+\sqrt{4 x^6 y^6+x^4 \epsilon  \left(16 y^4+\epsilon \right)+14 x^2 y^2 \epsilon ^2+y^4 \epsilon ^2}+y^2 \epsilon }{2 \left(x^2 y^2+\epsilon \right)^{3/2}}.
\end{align*}  


Taking the limits of $\lambda_{\epsilon,1}$ and $\lambda_{\epsilon,2}$ as $\epsilon \to 0$,
\begin{align*}
  \lambda_1 = \lim_{\epsilon \to 0} \lambda_{\epsilon, 1} &= \frac{4 a x^2 y^2 \sqrt{x^2 y^2}-2 \sqrt{x^6 y^6}}{2 \left(x^2 y^2\right)^{3/2}} \\
  &= 2 a - \frac{\left(x^2 y^2\right)^{3/2}}{\sqrt{x^6 y^6}} \\
  &= 2a - 1, \\
  \lambda_2 = \lim_{\epsilon \to 0} \lambda_{\epsilon, 2} &= \frac{4 a x^2 y^2 \sqrt{x^2 y^2}+2 \sqrt{x^6 y^6}}{2 \left(x^2 y^2\right)^{3/2}} \\
  &= 2 a+\frac{\left(x^2 y^2\right)^{3/2}}{\sqrt{x^6 y^6}} \\
  &= 2a + 1.
\end{align*}

In this form we see that $\nabla^2 f(x, y)$ has nonnegative eigenvalues if and only if $a \geq \frac{1}{2}$, and so $f$ is convex for $a \geq \frac{1}{2}$. To show strong convexity, we use the result that if matrix $M$ has eigenvalues $\left\{\lambda_i\right\}^n_{i = 1}$ then $M - k \mathbb I_n$ has eigenvalues $\left\{ \lambda_i - k \right\}^n_{i = 1}$. Therefore, $\nabla^2 f(x, y) - m \mathbb I_2$ has eigenvalues
\begin{align*}
  \lambda_{m, 1} &= 2a - 1 - m \\
  \lambda_{m, 2} &= 2a + 1 - m.
\end{align*}

To ensure $\lambda_{m, 1}, \lambda_{m, 2}$ are nonnegative we set $a > \frac{1}{2}$ and $m \leq a$. Therefore, $f$ is strongly convex with parameter $m$, $a \geq m > 0$, as desired.
\end{proof}

We present below figures of $f$ evaluated on $[-1, 1]\times [-1, 1]$ for $a \in \left\{0, 0.25, 0.5, 0.75\right\}$.

```{r, echo = F, fig.height = 4.5, fig.width = 3.5, fig.align = 'center', fig.show = 'hold'}
a <- c(0, 0.25, 0.5, 0.75)
f <- function(x, y, a) {
  abs(x * y) + a * (x^2 + y^2)
}

x <- y <- seq(-1, 1, length.out = 50)
Za <- lapply(a, function(ai) outer(x, y, FUN = function(x1, x2) f(x1, x2, ai)))

for (i in 1:length(a))
  persp(x, y, Za[[i]], zlab = expression(f(x, y)), theta = 15, phi = 30, 
        main = paste0("f(x, y; a), a = ", a[i]))
```

## 2.2

### 2.2 (a)

For $x \in \R^n_{++}$ we find gradient
$$
  \nabla f(x) = -\left[ x^{-1}_1, x^{-1}_2, \cdots , x^{-1}_n \right],
$$
and Hessian
$$
  \nabla^2 f(x) = 
  \begin{bmatrix}
    x^{-2}_1 & 0 & \cdots & 0 \\
    0 & x^{-2}_2 & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \cdots & x^{-2}_n \\
  \end{bmatrix}.
$$

Since $\nabla^2 f(x)$ is diagonal we may immediately obtain its eigenvalues $\left\{ \lambda_i \right\}^n_{i = 1}$,
$$
  \lambda_i = x^{-2}_i.
$$

We see that, since $x \in \R^n_{++} \iff x_i > 0, i = 1, ..., n$, all eigenvalues $\lambda_i > 0$. Therefore, $f$ must be strongly convex (and so strictly convex, and convex), as desired.

### 2.2 (b)

## 2.3

\begin{proof} ($\implies$) Suppose $f$ is convex. Then, $\text{dom}(f)$ is a convex set, and, for all $x, y \in \text{dom}(f)$ and $t \in [0, 1]$,
\begin{align*}
  f(tx + (1 - t)y) &\leq tf(x) + (1 - t)f(y) \\
  \iff f(t(x - y) + y) - f(y) &\leq t\left(f(x) - f(y)\right) \\
  \iff \frac{f(t(x - y) + y) - f(y)}{t} &\leq f(x) - f(y) \\
  \iff \frac{f(t(x - y) + y) - f(y)}{t} + f(y) &\leq f(x)
\end{align*}

Note that if we take the limit of our first term as $t\to 0$, for finite $x, y$,
\begin{align*}
  \lim_{t\to 0} \frac{f(t(x - y) + y) - f(y)}{t} &= \frac{\partial}{\partial t} f(t(x - y) + y)\Bigg]_{t = 0} \\
  &= \nabla f(t(x - y) + y)^T (x - y) \Bigg]_{t = 0} \\
  &= \nabla f(y)^T (x - y).
\end{align*}

Therefore, taking the limit of our inequality above as $t\to 0$,
\begin{align*}
  \lim_{t\to 0} \left( \frac{f(t(x - y) + y) - f(y)}{t} + f(y) \right) &\leq \lim_{t \to 0} f(x) \\
  \iff \nabla f(y)^T (x - y) + f(y) &\leq f(x).
\end{align*}

By symmetry we swap $x$ and $y$ to obtain
$$
  f \text{ convex} \implies \text{dom}(f) \text{ convex and } f(y) \geq f(x) + \nabla f(x)^T(y - x),
$$

as desired. \\

($\Longleftarrow$) Suppose $\text{dom}(f)$ is convex and, for $x, y \in \text{dom}(f)$, $x \neq y$,
$$
  \nabla f(x)^T (y - x) + f(x) \leq f(y).
$$

Since $\text{dom}(f)$ is convex we find $z = tx + (1 - t)y \in \text{dom}(f)$, $t \in [0,1]$. Then, for such $x, y, z$,
\begin{align*}
  \nabla f(z)^T (x - z) + f(z) &\leq f(x) \\
  \nabla f(z)^T (y - z) + f(z) &\leq f(y).
\end{align*}

Multiplying our first inequality by $t$ and the second by $(1 - t)$, and then adding the two yields
\begin{align*}
  t \left[ \nabla f(z)^T (x - z) + f(z) \right] + (1 - t) \left[ \nabla f(z)^T (y - z) + f(z) \right] &\leq tf(x) + (1 - t)f(y) \\
  \iff t\nabla f(z)^T(x - z) + (1 - t)\nabla f(z)^T(y - z) + f(z) &\leq tf(x) + (1 - t)f(y) \\ 
  \iff \nabla f(z)^T \left[ t(x - z) + (1 - t)(y - z) \right] + f(z) &\leq tf(x) + (1 - t)f(y) \\ 
  \iff \nabla f(z)^T \left[ tx + (1 - t)y - z \right] + f(z) &\leq tf(x) + (1 - t)f(y) \\ 
    \iff f(tx + (1 - t)y) &\leq tf(x) + (1 - t)f(y),
\end{align*}

where the final line was achieved by recalling that $z = tx + (1 - t)y$. Therefore,
$$
  f \text{ convex} \Longleftarrow \text{dom}(f) \text{ convex and } f(y) \geq f(x) + \nabla f(x)^T(y - x),
$$

as desired.
\end{proof}

# Question 3

## (a)

\begin{proof}

$(1 \implies 2)$
$(2 \implies 3)$
$(3 \implies 4)$
$(4 \implies 1)$

\end{proof}

## (b)

\begin{proof} Let $f$ be convex and twice differentiable. \\

$(1 \implies 2)$ If $\nabla f$ is $L$-Lipschitz then, for $x, y \in \text{dom}(f)$, $L > 0$,
$$
  \lVert \nabla^2 f(x) - \nabla^2 f(y) \rVert_2 \leq L \lVert x - y \rVert_2.
$$




$(2 \implies 3)$
$(3 \implies 4)$
$(4 \implies 1)$

\end{proof}

# Question 4

## (a)

### 1.

# Question 5