---
title: "MATH 680: Assignment 2"
author: "Annik Gougeon, David Fleischer"
date: "Last Update: `r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
header-includes:
   - \usepackage{amsmath,amsthm,amssymb,mathtools}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\R}{\mathbb R}
\renewcommand{\S}{\mathbb S}

# Question 1

\begin{proof} Let $C = \left\{ x \in \R^n: Ax \leq b \right\}$ be our set of interest. Let $x, y \in C$, and let $t \in [0, 1]$ be an arbitrary real-valued scalar. Then,
\begin{align*}
  A \left( tx + (1 - t)y \right) &= tAx + (1 - t)Ay \\
  &\leq tb + (1 - tb) \\
  &= b.
\end{align*}

Thus,
$$
  x, y \in C \implies tx + (1 - t)y \in C, \quad \text{for all } 0 \leq t \leq 1.
$$
That is, $C$ is a convex set, as desired.
\end{proof}

# Question 2

## 2.1

Let $f$ be defined as
$$
  f(x,y) = |xy| + a(x^2 + y^2).
$$
Recall that $f$ is a convex function if and only if its Hessian is positive semi-definite. 

**Case 1**: $xy > 0 \implies f_+(x,y) = xy + a(x^2 + y^2)$.

We begin by finding the gradient of $f_+(x,y)$ to be
$$
  \nabla f_+(x,y) = (y + 2ax, x + 2ay).
$$

Therefore, the Hessian is
$$
  \nabla^2 f_+(x,y)
  =
  \begin{bmatrix}
    2a & 1 \\
    1 & 2a
  \end{bmatrix}.
$$

For this Hessian to be positive semi-definite, we require its eigenvalues to be nonnegative. We find the eigenvalues to be $\lambda_1 = 2a - 1$ and $\lambda_2 = 2a + 1$. As a direct result, the Hessian is positive semi-definite (and therefore, convex) if and only if $a \geq \frac{1}{2}$.

Furthermore, in order for $f_+$ to be strongly convex, we must have the eigenvalues $\lambda \left(\nabla^2 f_+(x,y) - m \mathbb I \right)  = \{\lambda_i - m\}_{i=1,2}$ be nonnegative for some $m > 0$ (i.e., $\nabla^2 f_+(x,y) - m \mathbb I$ must be positive semidefinite for some $m > 0$). We see that
\begin{align*}
  \lambda_1 - m &= 2a - 1 - m \\
  \lambda_2 - m &= 2a + 1 - m.
\end{align*}

The above eigenvalues will be nonnegative for $a > \frac{1}{2}$ and $m \leq a$. Therefore, $f_+$ is strongly convex if $a > \frac{1}{2}$. 

**Case 2**: $xy < 0 \implies f_-(x,y) = -xy + a(x^2 + y^2)$.

We find $f_-$ to have gradient
$$
  \nabla f_-(x,y) = (-y +2ax , -x + 2ay),
$$

and Hessian
$$
  \nabla^2 f_-(x,y)
  =
    \begin{bmatrix}
    2a & -1 \\
    -1 & 2a
  \end{bmatrix}.
$$

Once again, for the Hessian to be positive semidefinite, we must find nonnegative eigenvalues $\lambda_i \geq 0$, $i = 1, 2$. As above, we find the eigenvalues to be $\lambda_1 = 2a - 1$ and $\lambda_2 = 2a + 1$. Therfore, the Hessian must be positive semidefinite (and so $f_-$ must be convex) if and only if $a \geq \frac{1}{2}$ and strongly convex if $a > \frac{1}{2}$.

**Case 3**: $xy = 0$. For such a case we note three possible scenarios
$$xy = 0 \implies  f_0(x,y) = 
\begin{cases} 
    ay^2, \quad \text{if } x = 0 \text{ and } y \ne 0, \\ 
    ax^2, \quad \text{if } x \ne 0 \text{ and } y = 0, \\
    0, \quad \quad \text{else.}
  \end{cases}
$$
    
If $x = 0$ and $y \ne 0$ then $f_0(x, y) =  ay^2$ so
$$
  \frac{\partial f_0}{\partial y} = 2ay,
$$

and
$$
  \frac{\partial^2 f_0}{\partial y^2} = 2a.
$$

It follows that for $x = 0$ and $y \ne 0$ we find $f_0(x,y)$ to be be convex for $a \geq \frac{1}{2}$ and strongly convex for $a > \frac{1}{2}$. By symmetry, the same result is found for the case where $x \ne 0$ and $y = 0$ (corresponding to $f_0(x, y) = ax^2$), permitting us to arrive to the same conclusions. 

Finally, for the last scenario $x = y = 0$, we note that the domain of $f_0$ under $x = y = 0$ is defined at only a single point, trivially satisfying strong convexity
$$
  \left( \nabla f(z_1) - \nabla f(z_2) \right)^T (z_1 - z_2) \geq m \lVert z_1 - z_2 \rVert^2_2
$$

on all points in its domain, as desired.

Presented below figures of $f$ evaluated on $[-1, 1]\times [-1, 1]$ for $a \in \left\{0, 0.25, 0.5, 0.75\right\}$.

```{r, echo = F, fig.height = 4.5, fig.width = 3.5, fig.align = 'center', fig.show = 'hold'}
a <- c(0, 0.25, 0.5, 0.75)
f <- function(x, y, a) {
  abs(x * y) + a * (x^2 + y^2)
}

x <- y <- seq(-1, 1, length.out = 50)
Za <- lapply(a, function(ai) outer(x, y, FUN = function(x1, x2) f(x1, x2, ai)))

for (i in 1:length(a))
  persp(x, y, Za[[i]], zlab = expression(f(x, y)), theta = 15, phi = 30, 
        main = paste0("f(x, y; a), a = ", a[i]))
```

## 2.2

### 2.2.a

For $x \in \R^n_{++}$ we find gradient of $f(x) = -\sum^n_{i = 1} \log x_i$ to be
$$
  \nabla f(x) = -\left[ x^{-1}_1, x^{-1}_2, \cdots , x^{-1}_n \right],
$$

with corresponding Hessian
$$
  \nabla^2 f(x) = 
  \begin{bmatrix}
    x^{-2}_1 & 0 & \cdots & 0 \\
    0 & x^{-2}_2 & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \cdots & x^{-2}_n \\
  \end{bmatrix}.
$$

Since $\nabla^2 f(x)$ is diagonal we may immediately obtain its eigenvalues $\left\{ \lambda_i \right\}^n_{i = 1}$ such that
$$
  \lambda_i = x^{-2}_i.
$$

We see that, since $x \in \R^n_{++} \iff x_i > 0, i = 1, ..., n$, all eigenvalues must be strictly positive $\lambda_i > 0$. Therefore, $f$ must be strongly convex (since there must exist constant $m$ such that $\lambda_i - m \geq 0$), as desired.

### 2.2.b

The entropy function, $f : \left\{x \in \mathbb R^n_+ : \sum_{i=1}^n x_i =1\right\} \to \R$ is defined as 
$$
f(x) = \begin{cases}
  -\sum_{i=1}^n x_i \log x_i & x > 0, \\
  0 & x = 0.
  \end{cases}
$$

Now, 
$$
  \nabla f(x) = \left(f'_1(x_1), f'_2(x_2), ..., f'_n(x_n)\right)
$$

where
$$
  f'_i(x_i) = 
  \begin{cases}
    -\left(1 + \log x_i\right), & x_i > 0 \\
    0 & x_i = 0.
  \end{cases}
$$

Likewise,
$$
  \nabla^2 f(x) = 
  \begin{bmatrix}
    f''_1(x_1) & 0 & \cdots & 0 \\
    0 & f''_2(x_2) & \cdots & 0 \\
    \vdots & \vdots & \ddots & \vdots \\
    0 & 0 & \cdots & f''_n(x_n)
  \end{bmatrix},
$$

where
$$
  f''_i(x_i) = 
  \begin{cases}
    -x^{-1}_i, & x_i > 0 \\
    0, & x_i = 0.
  \end{cases}
$$


Since $\nabla^2 f(x)$ is diagonal, we immediately obtain its eigenvalues 
$$
  \left\{ \lambda_i \right\}_{i = 1}^n = \left\{ f''_i(x_i) \right\}^n_{i = 1}
$$

Clearly, for any $x \in \text{dom}(f) = \R^n_+$ we find that all $\lambda_i \leq 0$. Since $\nabla^2 f$ always has nonpositive eigenvalues, we conclude that $\nabla^2 \left(-f\right)$ must have nonnegative eigenvalues. Therefore, $f$ is nonconvex (and in particular, $f$ is concave since $-f$ is convex), as desired.

## 2.3

\begin{proof} ($\implies$) Suppose $f$ is convex. Then, $\text{dom}(f)$ is a convex set, and, for all $x, y \in \text{dom}(f)$ and $t \in [0, 1]$,
\begin{align*}
  f(tx + (1 - t)y) &\leq tf(x) + (1 - t)f(y) \\
  &\iff f(t(x - y) + y) - f(y) \leq t\left(f(x) - f(y)\right) \\
  &\iff \frac{f(t(x - y) + y) - f(y)}{t} \leq f(x) - f(y) \\
  &\iff \frac{f(t(x - y) + y) - f(y)}{t} + f(y) \leq f(x)
\end{align*}

Note that if we take the limit of our first term as $t\to 0$, for finite $x, y$,
\begin{align*}
  \lim_{t\to 0} \frac{f(t(x - y) + y) - f(y)}{t} &= \frac{\partial}{\partial t} f(t(x - y) + y)\Bigg]_{t = 0} \\
  &= \nabla f(t(x - y) + y)^T (x - y) \Bigg]_{t = 0} \\
  &= \nabla f(y)^T (x - y).
\end{align*}

Therefore, taking the limit of our inequality above as $t\to 0$,
\begin{align*}
  \lim_{t\to 0} \left( \frac{f(t(x - y) + y) - f(y)}{t} + f(y) \right) &\leq \lim_{t \to 0} f(x) \\
  \iff \nabla f(y)^T (x - y) + f(y) &\leq f(x).
\end{align*}

By symmetry we may swap $x$ and $y$ to obtain
$$
  f \text{ convex} \implies \text{dom}(f) \text{ convex and } f(y) \geq f(x) + \nabla f(x)^T(y - x),
$$

as desired. \\

($\Longleftarrow$) Suppose $\text{dom}(f)$ is convex and, for $x, y \in \text{dom}(f)$, $x \neq y$,
$$
  \nabla f(x)^T (y - x) + f(x) \leq f(y).
$$

Since $\text{dom}(f)$ is convex we find $z = tx + (1 - t)y \in \text{dom}(f)$, $t \in [0,1]$. Then, for such $x, y, z$,
\begin{align*}
  \nabla f(z)^T (x - z) + f(z) &\leq f(x) \\
  \nabla f(z)^T (y - z) + f(z) &\leq f(y).
\end{align*}

Multiplying our first inequality by $t$ and the second by $(1 - t)$, and then adding the two yields
\begin{align*}
  t \left[ \nabla f(z)^T (x - z) + f(z) \right] &+ (1 - t) \left[ \nabla f(z)^T (y - z) + f(z) \right] \leq tf(x) + (1 - t)f(y) \\
  &\iff t\nabla f(z)^T(x - z) + (1 - t)\nabla f(z)^T(y - z) + f(z) \leq tf(x) + (1 - t)f(y) \\ 
  &\iff \nabla f(z)^T \left[ t(x - z) + (1 - t)(y - z) \right] + f(z) \leq tf(x) + (1 - t)f(y) \\ 
  &\iff \nabla f(z)^T \left[ tx + (1 - t)y - z \right] + f(z) \leq tf(x) + (1 - t)f(y) \\ 
    &\iff f(tx + (1 - t)y) \leq tf(x) + (1 - t)f(y),
\end{align*}

where the final line was achieved by recalling that $z = tx + (1 - t)y$. Therefore,
$$
  f \text{ convex} \Longleftarrow \text{dom}(f) \text{ convex and } f(y) \geq f(x) + \nabla f(x)^T(y - x),
$$

as desired.
\end{proof}

# Question 3

## 3.1

Let $f$ be convex and twice differentiable. 

### $(1 \implies 2)$ 

If $\nabla f$ is Lipschitz with constant $L > 0$ then, by definition,
$$
\lVert \nabla f(x) - \nabla f(y) \rVert_2 \leq L \lVert x - y \rVert_2.
$$

By the Cauchy-Schwartz inequality
$$
  \left| \langle u,v \rangle \right| \leq \lVert u\rVert \cdot \lVert v \rVert,
$$

substituting $u = \nabla f(x) - \nabla f(y)$ and $v = x - y$ in Cauchy-Schwartz, and by $L$-Lipschitz continuity assumed via (1), we obtain
\begin{align*}
\left( \nabla f(x) - \nabla f(y) \right)^T (x-y) &\leq \lVert \nabla f(x) - \nabla f(y) \rVert_2 \cdot \lVert x - y \rVert_2 \quad \text{(Cauchy-Schwartz)} \\
    &\leq L \lVert x - y \rVert^2_2 \quad \text{($L$-Lipschitz)},
\end{align*}

as desired. 

### $(1 \implies 3)$ 

Assume (1) holds
$$
  \lVert \nabla f(x) - \nabla f(y) \rVert_2 \leq L \lVert x - y \rVert_2,
$$

and define $g(t) = \nabla f\left(y + t(x - y)\right)$ so that
\begin{align*}
  g(0) &= \nabla f(y) \\
  g(1) &= \nabla f(x) \\
  g'(t) &= \nabla^2 f\left( y + t(x - y) \right) \cdot (x - y).
\end{align*}

Therefore,
$$
  \lVert \nabla f(x) - \nabla f(y) \rVert_2 = \lVert g(1) - g(0) \rVert_2.
$$

However, applying the Mean Value Theorem on $g$, there must exist some $z \in [0, 1]$ such that\footnote{We can be sure that such a point exists since we have assumed $f$ is twice differentiable (and so applying the Mean Value Theorem on $g$ is permissible) and we have assumed that $f$ is convex (so such a point $z \in [0,1]$ must exist).}
$$
  g'(z) = \frac{g(1) - g(0)}{1 - 0} = g(1) - g(0).
$$

Thus,
\begin{align*}
&\lVert \nabla f(x) - \nabla f(y) \rVert_2 \leq L \lVert x - y \rVert_2 \\
  &\implies \lVert g(1) - g(0) \rVert_2 \leq L \lVert x - y \rVert_2 \\
  &\implies \lVert g'(z) \rVert_2 \leq L \lVert x - y \rVert_2 \\
  &\iff \lVert \nabla^2 f\left( z^* \right) (x - y) \rVert_2 \leq L \lVert x - y \rVert_2, \quad \text{for some $z^* \in [x, y]$} \\
  &\iff \lVert \nabla^2 f\left(z^*\right) \rVert_2 \cdot \lVert x - y \rVert_2 \leq L \lVert x - y \rVert_2 \\
  &\iff \lVert \nabla^2 f\left(z^*\right) \rVert_2 \leq L \\
  &\iff \nabla^2 f\left(z^*\right) \preceq L \mathbb I.
\end{align*}

Therefore, since $x$ and $y$ were arbtirary
$$ 
  \nabla^2 f(x) \preceq L \quad \text{for all $x$},
$$

as desired. 

### $(3 \implies 4)$

Assume $(3)$ holds so that $\nabla^2 f(x) \preceq L\mathbb I$ and define $g$ such that
$$
  g(z) = \frac{L}{2} z^T z - f(z) \iff f(z) = \frac{L}{2} z^T z - g(z).
$$

Then,
$$
\nabla^2 f(x) \preceq L\mathbb I \iff 0 \preceq \nabla^2 g(x),
$$

informing us that $g$ must be convex. From the convexity of $g$, we must have that $g$ satisfies the first order condition outlined in 2.3
$$
  g(y) \geq g(x) + \nabla g(x)^T (y - x).
$$

Plugging in our definition of $g$ yields
\begin{align*}
  \frac{L}{2} y^T y - f(y) &\geq \frac{L}{2} x^T x - f(x) + \left(Lx - \nabla f(x)\right)^T\left(y - x\right) \\
  \iff -f(y) &\geq - f(x) - \nabla f(x)^T \left(y - x\right) + \frac{L}{2} x^T x - Lx^T x + Lx^T y - \frac{L}{2} y^T y \\
  &= - f(x) - \nabla f(x)^T \left(y - x\right) -\frac{L}{2} \left( y^T y - 2 x^T y + x^T x \right) \\
  &= -f(x) - \nabla f(x)^T\left(y - x\right) -\frac{L}{2} \lVert y - x \rVert^2_2 \\
  \iff f(y) &\leq f(x) + \nabla f(x)^T \left(y - x\right) + \frac{L}{2} \lVert y - x \rVert^2_2,
\end{align*}

as desired.

### $(4 \implies 1)$ 

Assume (4) holds so that
$$
  f(y) \leq f(x) + \nabla f(x)^T(y - x) + \frac{L}{2} \lVert y - x \rVert^2_2.
$$

Next, define $g(z) = \frac{L}{2} z^T z - f(z) \iff f(z) = \frac{L}{2}z^Tz - g(z)$ and apply (4) to yield
\begin{align*}
  \frac{L}{2}y^T y - g(y) &\leq \frac{L}{2} x^T x - g(x) + \left(Lx - \nabla g(x)\right)^T \left(y - x\right) + \frac{L}{2} \lVert y - x \rVert^2_2 \\
  \iff g(x) - g(y) &\leq -\frac{L}{2} x^T x - \frac{L}{2} y^T y + L x^T y - L x^T x - \nabla g(x)^T (y - x) + \frac{L}{2} \lVert y - x \rVert^2_2 \\
  &= -\frac{L}{2} \left( y^T y - 2 x^T y + x^T x \right) - \nabla g(x)^T(y - x) + \frac{L}{2} \lVert y - x \rVert^2_2 \\
  &= -\frac{L}{2} \lVert y - x \rVert^2_2 - \nabla g(x)^T (y - x) + \frac{L}{2} \lVert y - x \rVert^2_2 \\
  &= -\nabla g(x)^T(y - x) \\
  \iff g(y) & \geq g(x) + \nabla g(x)^T(y - x).
\end{align*}

Thus, recalling our first order condition for convexity from Question 2.3 we may conclude that $g$ is indeed convex. Now, if $g$ is convex then
$$
  \nabla^2 g(x) \succeq 0 \iff LI \succeq \nabla^2 f(x).
$$

However, defining $h(t) = \nabla f(y + t(x - y))$ so that
\begin{align*}
  h(0) &= \nabla f(x) \\
  h(1) &= \nabla f(y) \\
  h'(t) &= \nabla^2 f(y + t(x - y)) \cdot (x - y).
\end{align*}

Therefore,
\begin{align*}
  \lVert \nabla f(x) - \nabla f(y) \rVert_2 &= \lVert \nabla f(y) - \nabla f(x) \rVert_2 \\
  &= \lVert h(1) - h(0) \rVert_2 \\
  &= \| \int^1_0 h'(t) \,dt \|_2 \\
  &\leq \int^1_0 \lVert h'(t) \rVert_2\,dt \quad \text{(triangle inequality)} \\
  &= \int^1_0 \lVert \nabla^2 f\left( y + t(x - y) \right) \cdot (x - y) \rVert_2\,dt \\
  &\leq \int^1_0 L \lVert x - y \rVert_2\,dt \\
  &= L \lVert x - y \rVert,
\end{align*}

where the final inequality was obtained through
$$
  \nabla^2 f(x) \preceq L\mathbb I \iff \lVert \nabla^2 f(x) \rVert_2 \leq L.
$$

Putting this together we find that $f(y) \leq f(x) + \nabla f(x)^T(y - x) + \frac{L}{2} \lVert y - x \rVert^2_2$ implies that $\nabla f$ is $L$-Lipschitz, as desired. \\

Since we have shown $(1) \implies (2), (3)$, $(3) \implies (4)$, and $(4) \implies (1)$ we may conclude that all the above statements are indeed equivalent.

## 3.2

### $(1 \implies 2)$

Assume (1) so that $f$ is strongly convex with parameter $m > 0$. Then, by definition, $f(x) - \frac{m}{2} \lVert x \rVert^2_2$ is convex. Defining $g(x) = f(x) - \frac{m}{2} \lVert x \rVert^2_2$ we find
$$
  \nabla g(x) = \nabla f(x) - m x.
$$


However, recall that $g$ is convex if and only if 
$$
  0 \leq \left( \nabla g(x) - \nabla g(y) \right)^T (x - y).
$$

Therefore,
\begin{align*}
  0 &\leq \left( \nabla f(x) - mx - \nabla f(y) + my \right)^T(x - y) \\
  &= \left(\nabla f(x) - \nabla f(x)\right)^T(x - y) - m(x - y)^T(x - y) \\
  &= \left(\nabla f(x) - \nabla f(x)\right)^T(x - y) - m\lVert x - y \rVert^2_2 \\
  \iff m \lVert x - y \rVert^2_2 &\leq \left(\nabla f(x) - \nabla f(x)\right)^T(x - y),
\end{align*}
as desired.

### $(2 \implies 3)$

Assuming (2) holds we find
\begin{align*}
&\left(\nabla f(x) - \nabla f(y) \right)^T (x-y) \geq m\lVert x-y \rVert^2_2 \\
    &\iff \nabla f(x) + x\nabla^2 f(x) -y \nabla^2 f(x) + y\nabla f(y) \geq 2m\lVert x-y \rVert \\
    &\iff \nabla^2 f(x) (x-y) + \left(\nabla f(x) - \nabla f(y)\right) \geq 2m\lVert x-y \rVert \\
    &\iff \nabla^2 f(x) \lVert x-y \rVert \geq m \lVert x-y \rVert,
\end{align*}

since $\lVert \nabla f(x) - \nabla f(y) \rVert \geq m \lVert x-y \rVert$. It follows that
$$
\nabla^2 f(x) \succeq mI,
$$
as desired. 

### $(1 \implies 4)$

Once again assume (1) so that $f$ is strongly convex with parameter $m > 0$. Then, by definition, $f(x) - \frac{m}{2} \lVert x \rVert^2_2$ is convex. Defining $g(x) = f(x) - \frac{m}{2} \lVert x \rVert^2_2$ we find
$$
  \nabla g(x) = \nabla f(x) - m x.
$$

Now, apply the first order condition from 2.3 on $g$
$$
  g(y) \geq g(x) + \nabla g(x)^T (y - x).
$$

Hence,
\begin{align*}
  f(y) - \frac{m}{2} \lVert y \rVert^2_2 &\geq f(x) - \frac{m}{2} \lVert x \rVert^2_2 + \left( \nabla f(x) - mx \right)^T (y - x) \\
  \iff f(y) &\geq f(x) + \nabla f(x)^T (y - x) + \frac{m}{2} y^T y - \frac{m}{2} x^T x - mx^T y + mx^T x \\
  &= f(x) + \nabla f(x)^T(y - x) + \frac{m}{2} y^T y - mx^Ty + \frac{m}{2} x^T x \\
  &= f(x) + \nabla f(x)^T(y - x) + \frac{m}{2} \lVert y - x \rVert^2_2 \\
  \iff f(y) &\geq f(x) + \nabla f(x)^T (y - x) + \frac{m}{2} \lVert y - x \rVert^2_2,
\end{align*}

as desired.

### $(4 \implies 1)$

We begin with 
\begin{align*}
&f(y) \geq f(x) + \nabla f(x)^T (y-x) + \frac{m}{2} \lVert y-x \rVert^2_2 \\
    &\implies \nabla f(y) \geq  \nabla f(x) + m\lVert y-x\rVert \\
    &\implies \lVert \nabla f(y) - \nabla f(x) \rVert \geq m\lVert y-x \rVert \\ 
    &\implies \lVert \nabla f(x) - \nabla f(y) \rVert \geq m\lVert x-y \rVert. 
\end{align*}

Therefore, $f$ must be strongly convex, as desired. 

# Question 4

## 4.a

For parts 1 and 2 we load `CVXR` in order to solve the (convex) 2d fused lasso problem.
```{r, message = F}
library(CVXR)
```

### 4.a.1

We load our data `circle.csv` and define some useful constants
```{r}
circle <- as.matrix(read.csv("../data/circle.csv", header = F))
n <- length(circle); nr <- nrow(circle); nc <- ncol(circle)
```

Next, we translate the 2d fused lasso penalty
$$
  \lambda \sum_{\{i,j\} \in E} |\theta_i - \theta_j|, \quad E = \text{set of edges \{(i,j)\} connecting adjascent pixels}
$$

into a `CVXR` compatible function
```{r}
fused_lasso_2d <- function(theta, lambda = 0) {
  nr <- nrow(theta); nc <- ncol(theta)
  S <- theta[1:(nr - 1),] - theta[2:nr,] # SOUTH
  N <- theta[2:nr,] - theta[1:(nr - 1),] # NORTH
  E <- theta[,1:(nc - 1)] - theta[,2:nc] # EAST
  W <- theta[,2:nc] - theta[,1:(nc - 1)] # WEST
  lambda * (sum(abs(S)) + sum(abs(N)) + sum(abs(E)) + sum(abs(W)))
}
```

as well as defining some parameters and variables, as well as our $\ell_2$ loss $\frac{1}{2} \lVert Y - \theta \rVert^2_2$ 
```{r}
lambda <- 1
theta <- Variable(nr, nc)
theta_hat <- matrix(0, nrow = nr, ncol = nc)
loss <- sum(0.5 * (circle - theta)^2)
```

Finally, we run run `CVXR` on the problem
```{r}
obj <- loss + fused_lasso_2d(theta, lambda)
prob <- Problem(Minimize(obj))
res <- solve(prob)
theta_hat <- res$getValue(theta)
```

Comparing the original data with the fused fit we see that the fused fit of the circle is essentially a square with truncated/rounded corners, as seen below.
```{r, echo = F, fig.show = 'hold', fig.align = 'center', fig.width = 3.5, fig.height = 4}
cols <- colorRampPalette(c("black", "white"))
image(circle, xaxt = 'n', yaxt = 'n', col = cols(16), main = "Target")
image(theta_hat, xaxt = 'n', yaxt = 'n', col = cols(16),
      main = substitute(paste("Fused Lasso Fit: ", lambda, " = ", lam),
                        list(lam = lambda)))
```

We can understand such behaviour by the noting the conflicting behaviour of the $\ell_2$ term and the $\ell_1$ fused term in our objective function. The $\ell_2$-loss term
$$
  \frac{1}{2}\sum^n_{i = 1} \left(y_i - \theta_i\right)^2
$$

is minimized as $\hat\theta_i \longrightarrow y_i$, while the $\ell_1$-fused term
$$
  \lambda \sum_{\{i,j\} \in E} |\theta_i - \theta_j|
$$

is minimized when adjascent cells are 'close' to each other (and as $\lambda \to \infty$ we obtain $\hat\theta_i \to \bar y$). Deep in the interior/far out in the extertior of the circle, the objective estimates $\hat\theta_i \approx y_i$ since essentially all adjascent values will correspond to 0 (black) or 1 (white), respectively. However, along the boundary of the two regions we balance the two objectives by detecting a 'changepoint' in the data (as a consequence of using the $\ell_1$ norm), inside of which most observations are close to 0 and beyond which most observations are close to 1.

### 4.a.2

We load the `lenna_64.csv` data and define some useful constants
```{r}
lenna <- as.matrix(read.csv("../data/lenna_64.csv", header = F))
n <- length(lenna); nr <- nrow(lenna); nc <- ncol(lenna)
```

and run `CVXR` on the 2d fused lasso problem in the same way as we did for `circle.csv`, but now over a vector of tuning parameters $\{\lambda_k\}_{k = 0, ..., 8} = \left\{10^{-k/4}\right\}_{k = 0, ..., 8}$

```{r}
lambda_vals <- 10^(-(0:8)/4)
theta_vals <- vector(mode = 'list', length = length(lambda_vals))
obj_val <- vector(mode = 'numeric', length = length(lambda_vals))
theta <- Variable(nr, nc)
loss <- sum(0.5 * (lenna - theta)^2)

pt <- proc.time()
for (i in 1:length(lambda_vals)) {
  lambda <- lambda_vals[i]
  obj <- loss + fused_lasso_2d(theta, lambda)
  prob <- Problem(Minimize(obj))
  res <- solve(prob)
  theta_vals[[i]] <- res$getValue(theta)
  obj_val[i] <- res$value
}
print(proc.time() - pt)
```

Presented below are plots of the original data and the fused fits
```{r, echo = F, fig.align = 'center', fig.height = 4, fig.width = 4}
cols <- colorRampPalette(c("black", "white"))
# flip so image() diplays the right orientation
lenna2 <- t(apply(lenna, 2, rev))
theta_hat <- lapply(theta_vals, function(tv) t(apply(tv, 2, rev)))

image(lenna2, col = cols(16), xaxt = 'n', yaxt = 'n', main = "Target")
hist(lenna2, breaks = 1e2, xlim = c(0, 1), main = "Target",
     xlab = expression(hat(theta)))
```

```{r, echo = F, fig.align = 'center', fig.height = 3.25, fig.width = 3, fig.show = 'hold'}
for (k in 1:length(lambda_vals)) {
  image(theta_hat[[k]], col = cols(16), xaxt = 'n', yaxt = 'n',
        main = substitute(paste(lambda, " = ", lam),
                          list(lam = round(lambda_vals[k], 3))))
}
```

along with histograms of the fit values $\{\hat\theta_i\}$
```{r, echo = F, fig.show = 'hide'}
# find largest count in histograms in order to standardize
# the histograms to the same y axis
ymax <- max(sapply(theta_hat, function(th) max(hist(th)$counts)))
```

```{r, echo = F, fig.align = 'center', fig.height = 3, fig.width = 2.75, fig.show = 'hold'}
for (k in 1:length(lambda_vals)) {
  hist(theta_hat[[k]], breaks = 1e2, xlim = c(0, 1),
       ylim = c(0, ymax),
       xlab = expression(hat(theta)),
       ylab = "",
       main = substitute(paste(lambda, " = ", lam),
                        list(lam = round(lambda_vals[k], 3))))
}
```

Note that as $\lambda \to 0$ we find the distribution of $\hat\theta_i$ generally becomes less and less kurt. That is, when $\lambda$ is large we see nearly all the fits to be near the mean $\bar y$, and becoming more and more dispersed (towards $\hat\theta_i \to y_i$) as $\lambda$ shrinks.

## 4.b

### 4.b.1

Note that our expression
$$
  \lVert (x, y) \rVert^3_1 \leq 5x + 7
$$

is successfully recognized as convex without any serious manipulations. That is, in DCP we write this as
$$
  \lVert (x, y) \rVert^3_1 \leq 5x + 7 \quad \mapsto \quad \texttt{pow(norm1(x, y), 3) <= 5 * x + 7},
$$

as desired.

### 4.b.2

We now consider the expression
$$
  \frac{2}{x} + \frac{9}{z - y} \leq 3.
$$

Since DCP automatically constraining the argument to be within the function's domain handling the domain, DCP does not allow division $\frac{a}{b}$ to be input as `a/b`, as we may expect. Instead, DCP accepts `inv_pos(x)` as $\frac{1}{x}$ and restricts $x$ to $x > 0$ to enforce convexity. Therefore, we replace any instance of $\frac{1}{x}$ with `inv_pos(x)` to yield

$$
  \frac{2}{x} + \frac{9}{z - y} \leq 3 \quad \mapsto \quad \texttt{2 * inv\_pos(x) + 9 * inv\_pos(z - y) <= 3}
$$

as desired.

### 4.b.3

We have the expression
$$
  \sqrt{x^2 + 4} + 2y \leq -5x.
$$

The RHS is valid as-is, while the LHS requires some manipulation in order for the problem to be recognized as convex since $\sqrt{\cdot}$ is treated as concave, independent of its arguments. To get around this we make use of the $\ell_2$ norm (which is classified as convex)
$$
  \lVert (x_1, x_2) \rVert_2 = \sqrt{x_1^2 + x_2^2}.
$$

Hence,
$$
  \lVert (x, 2) \rVert_2 = \sqrt{x^2 + 4}.
$$

This gives us the following DCP expressions
$$
  \sqrt{x^2 + 4} + 2y \leq -5x \quad \mapsto \quad \texttt{norm2(x, 2) + 2 * y <= -5 * x}
$$

as desired.

### 4.b.3

We begin with the problem
$$
  (x + 3) \cdot z \cdot (y - 5) \geq 8,\quad x \geq -3,\,z \geq 0,\,y \geq 5.
$$

To translate into DCP we first note that this is equivalent to the problem
$$
  x\cdot z \cdot y \geq 8,\quad x \geq 0,\,z \geq 0,\,y \geq 0.
$$

To enforce the (new) domain $x, y, z \geq 0$ we apply the geometric mean
$$
  \texttt{geo\_mean(x1, ..., xk)} = \left(x_1 \cdot \cdots \cdot x_k \right)^\frac{1}{k}
$$

since DCP automatically restricts each argument as $x_i \geq 0$. We now wish to remove the $k^\text{th}$-root term. However, if we were to raise our geometric mean expression to the $k^\text{th}$ power then DCP would no longer treat the LHS as a concave expression since $x^k$, $k > 1$, is always considered to be convex (independent of the form $x$ takes). A solution is to instead take the $k^\text{th}$ root of the RHS. This gives us the DCP expression
$$
  (x + 3) \cdot z \cdot (y - 5) \geq 8,\quad x \geq -3,\,z \geq 0,\,y \geq 5 \quad \mapsto \quad \texttt{geo\_mean(x, y, z) >= 2},
$$

as desired.

### 4.b.5

Our expression
$$
  \frac{\left(x + 3z\right)^2}{\log \left(y - 1\right)} + 2 y^2 \leq 10
$$

is translated into DCP by making use of the function $\frac{s^2}{t} \mapsto \texttt{quad\_over\_lin(s, t)}$. Without other major issues we translate this expression directly into DCP via
$$
  \frac{\left(x + 3z\right)^2}{\log \left(y - 1\right)} + 2 y^2 \leq 10 \quad \mapsto \quad \texttt{quad\_over\_lin(x + 3 * z, log(y - 1)) + 2 * square(y) <= 10},
$$

as desired.

### 4.b.6

We wish to translate the following into DCP interpretable format
$$
    \log \left( e^{-\sqrt{x}} + e^{2z} \right) \leq -e^{5y}.
$$

DCP has a unique function designed to handle logarithms of sums of exponential terms. In paritcular,
$$
  \log \left( e^{x_1} + \cdots + e^{x_k} \right) \quad \mapsto \quad \texttt{log\_sum\_exp(x1, ..., xk)}.
$$

Using this scheme we translate our expression into DCP as
$$
  \log \left( e^{-\sqrt{x}} + e^{2z} \right) \leq -e^{5y} \quad \mapsto \quad \texttt{log\_sum\_exp(-sqrt(x), 2 * z) <= -exp(5 * y)},
$$

as desired.

### 4.b.7

We begin by noting the string of equivalences of our target expression
\begin{align*}
\sqrt{\lVert \left(2x - 3y, y + x\right) \rVert_1} = 0 &\iff \lVert \left(2x - 3y, y + x\right) \rVert_1 = 0 \\
&\iff \lVert \left(2x - 3y, y + x\right) \rVert_1 \leq 0.
\end{align*}

This yields the DCP expression
$$
  \sqrt{\lVert \left(2x - 3y, y + x\right) \rVert_1} = 0 \quad \mapsto \quad \texttt{norm1(2 * x - 3 * y, y + x) <= 0}.
$$

as desired.

### 4.b.8

We wish to translate the following inequality
$$
  y \log \left (\frac{y}{2x}\right) \leq y + x, \quad x > 0,\, y > 0.
$$

DCP handles the LHS via the Kullback–Leibler function $y \log \frac{y}{x} - y + x \mapsto \texttt{kl\_div(y, x)}$. This function automatically handles the domain restriction of $x, y > 0$. Thus,
$$
  y \log \left (\frac{y}{2x}\right) \leq y + x, \quad x > 0,\, y > 0 \quad \mapsto \quad \texttt{kl\_div(y, 2 * x) + x - y <= y + x - 30},
$$

as desired. 



\newpage

# Appendix

## Question 2.1 (Alternate Solution)

Below we present an alternate (and terrible) solution to proving that $f(x, y) = |xy| + a\left(x^2 + y^2\right)$ is convex $\iff a \geq \frac{1}{2}$ and strictly convex $\iff a > \frac{1}{2}$.

\begin{proof} Recall that a (continuous, twice differentiable) function $f(z)$, $z\in C$, is convex on $C$ if and only if its Hessian is positive semidefinite for all $z$ on the interior of $C$,
$$
  \nabla^2 f(z) \in \S^n_+,
$$
and strongly convex with parameter $m > 0$ if and only if
$$
  \nabla^2 f(z) - m \mathbb I_n \in \S^n_{+}.
$$

Furthermore, a matrix $M \in \R^{n\times n}$ is positive semidefinite if and only if all eigenvalues of $M$ are nonnegative. Since $f$ is nondifferentiable along $x = 0, y = 0$ we first apply a differentiable approximation $f_\epsilon$
$$
  f_\epsilon(x, y) = \sqrt{x^2 y^2 + \epsilon} + a\left(x^2 + y^2\right) \underset{\epsilon\to 0}{\longrightarrow} |xy| + a\left(x^2 + y^2\right) = f(x, y)
$$
Now, $f_\epsilon$ admits gradient 
$$
  \nabla f_\epsilon(x, y) = \left(2 a x+\frac{x y^2}{\sqrt{x^2 y^2+\epsilon }},2 a y+\frac{x^2 y}{\sqrt{x^2 y^2+\epsilon }}\right),
$$

and Hessian
$$
  \nabla^2 f_\epsilon(x, y) = \left[
\begin{array}{cc}
 -\frac{x^2 y^4}{\left(x^2 y^2+\epsilon \right)^{3/2}}+\frac{y^2}{\sqrt{x^2 y^2+\epsilon }}+2 a & \frac{2 x y}{\sqrt{x^2 y^2+\epsilon }}-\frac{x^3 y^3}{\left(x^2 y^2+\epsilon \right)^{3/2}} \\
 \frac{2 x y}{\sqrt{x^2 y^2+\epsilon }}-\frac{x^3 y^3}{\left(x^2 y^2+\epsilon \right)^{3/2}} & -\frac{y^2 x^4}{\left(x^2 y^2+\epsilon \right)^{3/2}}+\frac{x^2}{\sqrt{x^2 y^2+\epsilon }}+2 a \\
\end{array}
\right].
$$

We find $\nabla^2 f_\epsilon(x, y)$ to have eigenvalues\footnote{Details left as an exercise.}
\begin{align*}
  \lambda_{\epsilon, 1} &= \frac{x^2 \left(4 a y^2 \sqrt{x^2 y^2+\epsilon }+\epsilon \right)+4 a \epsilon  \sqrt{x^2 y^2+\epsilon }-\sqrt{4 x^6 y^6+x^4 \epsilon  \left(16 y^4+\epsilon \right)+14 x^2 y^2 \epsilon ^2+y^4 \epsilon ^2}+y^2 \epsilon }{2 \left(x^2 y^2+\epsilon \right)^{3/2}} \\
  \lambda_{\epsilon, 2} &= \frac{x^2 \left(4 a y^2 \sqrt{x^2 y^2+\epsilon }+\epsilon \right)+4 a \epsilon  \sqrt{x^2 y^2+\epsilon }+\sqrt{4 x^6 y^6+x^4 \epsilon  \left(16 y^4+\epsilon \right)+14 x^2 y^2 \epsilon ^2+y^4 \epsilon ^2}+y^2 \epsilon }{2 \left(x^2 y^2+\epsilon \right)^{3/2}}.
\end{align*}  


Taking the limits of $\lambda_{\epsilon,1}$ and $\lambda_{\epsilon,2}$ as $\epsilon \to 0$,
\begin{align*}
  \lambda_1 = \lim_{\epsilon \to 0} \lambda_{\epsilon, 1} &= \frac{4 a x^2 y^2 \sqrt{x^2 y^2}-2 \sqrt{x^6 y^6}}{2 \left(x^2 y^2\right)^{3/2}} \\
  &= 2 a - \frac{\left(x^2 y^2\right)^{3/2}}{\sqrt{x^6 y^6}} \\
  &= 2a - 1, \\
  \lambda_2 = \lim_{\epsilon \to 0} \lambda_{\epsilon, 2} &= \frac{4 a x^2 y^2 \sqrt{x^2 y^2}+2 \sqrt{x^6 y^6}}{2 \left(x^2 y^2\right)^{3/2}} \\
  &= 2 a+\frac{\left(x^2 y^2\right)^{3/2}}{\sqrt{x^6 y^6}} \\
  &= 2a + 1.
\end{align*}

In this form we see that $\nabla^2 f(x, y)$ has nonnegative eigenvalues if and only if $a \geq \frac{1}{2}$, and so $f$ is convex for $a \geq \frac{1}{2}$. To show strong convexity, we use the result that if matrix $M$ has eigenvalues $\left\{\lambda_i\right\}^n_{i = 1}$ then $M - k \mathbb I_n$ has eigenvalues $\left\{ \lambda_i - k \right\}^n_{i = 1}$. Therefore, $\nabla^2 f(x, y) - m \mathbb I_2$ has eigenvalues
\begin{align*}
  \lambda_{m, 1} &= 2a - 1 - m \\
  \lambda_{m, 2} &= 2a + 1 - m.
\end{align*}

To ensure $\lambda_{m, 1}, \lambda_{m, 2}$ are nonnegative we set $a > \frac{1}{2}$ and $m \leq a$. Therefore, $f$ is strongly convex with parameter $m$, $a \geq m > 0$, as desired.
\end{proof}
